---
title: "An Exploratory Analysis of Wearable Fitness Step Tracking Data"
output: html_document
---
#Introduction

This document is a summary of some exploratory analysis of wearable fitness data that tracks steps by day in 5 minute increments.  The dplyr package was used to perform most of the analysis and transformations as required.

After setting the working directory to the project directory, the first task was reading in the data into a data frame named fitdata.  This was accomplished with the following R code:

```{r, echo=TRUE}
fitdata <- read.csv("activity.csv")

```

Once the data is available, I chose to use the dplyr package to perform the transformations and summary data analysis.  A nice dplyr summary can be found [here.](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html)  I used dplyr to create three data frames:
* **fitdatabydate** which sums steps by date
* **fitdatabyinterval** which averages the number of steps taken in each observed interval
* **fitdata_nafill** which uses the **fitdatabyinterval** average to fill in any interval which had NA recorded in the steps feature.

Here is the code creating the three data frames:

```{r, echo = TRUE}

require(dplyr)
# Sum up the steps taken per day
fitdatabydate <- summarise(group_by(fitdata, date), tot_steps = sum(steps, na.rm = TRUE)) 

# Calculate the mean number of steps taken during each interval in the data set
fitdatabyinterval <- summarise(group_by(fitdata, interval), mean_steps = mean(steps, na.rm = TRUE)) 

#join the interval mean data set to the original fitness data
fitdata_nafill <- left_join(fitdata, fitdatabyinterval, by = "interval")

#replace any NA values with the mean value for that interval
fitdata_nafill$steps[is.na(fitdata_nafill$steps)] <- fitdata_nafill$mean_steps

```


Now that the data has been transformed, some exploratory analysis can be conducted.  Here is a histogram of the total number of steps taken per day

```{r, echo=TRUE}
hist(fitdatabydate$tot_steps, breaks = 30, col = "blue", xlab = "Steps Per Day", ylab = "Number of days", main = "Histogram of Recorded Steps per Day")

```

Calculating the mean and median of steps per day, use:

```{r}

mean(fitdatabydate$tot_steps, na.rm = TRUE)
median(fitdatabydate$tot_steps, na.rm = TRUE)

```

#Analyzing Average Daily Patterns

To analyze the average daily patterns, I created the **fitdatabyinterval** dataset.  Using the following code, the time series for the data set is:

```{r, echo=TRUE}

plot(x = fitdatabyinterval$interval, y = fitdatabyinterval$mean_steps, type = "l", main = "Average Number of Steps Per Interval (All Days Combined)", xlab = "Interval", ylab = "Average Steps")

```


#Imputing Missing Values

A quick check of the data reveals there are numerous NAs in the data.
```{r, echo=TRUE}
num_nas <- sum(is.na(fitdata$steps))

```

I chose to replace the NA values with the mean for that specific interval.  I did this by using the merging the summary dataframe fitdatabyinterval with the fitdata detailed dataframe.  Using the imputed data, we can now compare it to the source data, first with a histogram and then by comparing the mean and median values of steps per day as well as total steps per day

```{r, echo=TRUE}
fitdata_nafill <- left_join(fitdata, fitdatabyinterval, by = "interval")
fitdata_nafill$steps[is.na(fitdata_nafill$steps)] <- fitdata_nafill$mean_steps

#Generate a histogram of the dataframe with imputed values
fitdata_imputed_bydate <- summarize(group_by(fitdata_nafill, date), tot_steps = sum(steps))
hist(fitdata_imputed_bydate$tot_steps, breaks = 30, col = "red", xlab = "Steps Per Day", ylab = "Number of days", main = "Histogram of Recorded Steps per Day (Imputed NAs)")

ImputedChange_dailymean <- mean(fitdata_imputed_bydate$tot_steps) - mean(fitdatabydate$tot_steps, na.rm = TRUE)
ImputedChange_dailymedian <- median(fitdata_imputed_bydate$tot_steps) - median(fitdatabydate$tot_steps, na.rm = TRUE)

#Create a data frame to compare daily changes in step totals after imputing values
ImputedChange_dailytotalsteps <- inner_join(fitdatabydate,fitdata_imputed_bydate,by="date")

```

#Compare Weekdays to Weekends

Lastly, compare weekday to weekend patterns.  There is a noticeable difference in the latter part of the day with higher activity associated in the afternoon on weekends.

```{r, echo=TRUE}
fitdata_nafill$isweekend <- weekdays(as.Date(fitdata_nafill$date)) %in% c("Saturday","Sunday")
fitdata_nafill$isweekend[weekdays(as.Date(fitdata_nafill$date)) %in% c("Saturday","Sunday")] <- as.factor("Weekend")
fitdata_nafill$isweekend[fitdata_nafill$isweekend==0] <- "weekday"
fitdata_nafill$isweekend[fitdata_nafill$isweekend==TRUE] <- "weekend"
df_weekdays <- filter(fitdata_nafill, isweekend == "weekday")
df_weekdays <- summarise(group_by(df_weekdays, interval), mean_steps = mean(steps))
df_weekends <- filter(fitdata_nafill, isweekend == "weekend")
df_weekends <- summarise(group_by(df_weekends, interval), mean_steps = mean(steps))
par(mfrow=c(2,1))
plot(x = df_weekdays$interval, y = df_weekdays$mean_steps, type = "l", main = "Average Steps Per Interval (Weekdays)", xlim = c(0,2500), ylim = c(0,300))
plot(x = df_weekends$interval, y = df_weekends$mean_steps, type = "l", main = "Average Steps Per Interval (Weekend)", xlim = c(0,2500), ylim = c(0,300))


```

